Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population

### 2025-12-02 manuscript/manuscript.qmd
Update manuscript

---
title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
date: "2025-11-25"
author:
  - name: Cormac Monaghan
    url: https://c-monaghan.github.io/
    email: cormacmonaghan@protonmail.com
    orcid: 0000-0001-9012-3060
    affiliation:
      - id: 1
        name: Maynooth University
    attributes:
      corresponding: true
  - name: Rafael de Andrade Moral
    url: https://rafamoral.github.io/
    orcid: 0000-0002-0875-3563
    affiliation:
      - id: 1
        name: Maynooth University
  - name: Michelle Kelly
    url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
    orcid: 0000-0001-6159-6707
    affiliation:
      id: 2
      name: National College of Ireland
  - name: Joanna McHugh Power
    url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
    orcid: 0000-0002-7387-3107
    affiliation:
      id: 1
      name: Maynooth University
correspondence-prefix: "Corresponding author"
abstract: |
  Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential. While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction. We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} = 69.70;\sigma = 7.58)$. Cognitive function, procrastination, and a proxy measure of apathy were assessed. Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model. Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function. Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants. Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/C-Monaghan/crisp-clam>
keywords:
  - Dementia risk
  - Mild cognitive impairment
  - Procrastination
  - Apathy
  - Executive dysfunction
  - Aging
published: "This manuscript is a preprint and is currently under review at *Alzheimer's & Dementia: Diagnosis, Assessment & Disease Monitoring*"
code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
bibliography: references.bib
csl: csl/apa.csl
block-author: false
link-citations: true
editor:
  markdown:
    wrap: 80
keep-tex: true
---

@@

The Lancet Commission on Dementia identified fourteen causal risk factors
including smoking, hypertension, and diabetes, which account for approximately
45% of dementia cases worldwide [@livingston2024]. Prodromal markers, on the
other hand, reflect early disease processes and may signal limited opportunities
for prevention [@teipel2025]. One such prodromal marker is apathy, or the
significant loss of motivation, as distinct from depression and cognitive
impairment [@fresnais2023; @richard2012]. Apathy is prevalent in both MCI and
dementia subgroups [@donovan2014; @fresnais2023; @salem2023; @vandalen2018].
Apathy is a significant predictor of the transition from MCI to dementia
[@ruthirakuhan2019; @salem2023; @vandalen2018]. While apathy has been
well-established as a prodromal marker and risk factor for dementia progression,
emerging evidence suggests that procrastination may also relate to early
cognitive decline.

Although apathy and procrastination can appear similar in that both lead to
reduced engagement in everyday activities, the underlying behavioural processes
are distinct. Apathy reflects reduced internal drive and emotional engagement,
impairing initiation of action [@fresnais2023]. In contrast, procrastination is
a self-regulatory failure defined by the voluntary delay of an important and
intended course of action despite expecting negative consequences [@sirois2013;
@klingsieck2013]. This distinction suggests that individuals who procrastinate
typically form intentions but fail to execute them in a timely manner. This
failure can often be understood through an affective framework, where delayed
action may stem from the avoidance of negative emotions (i.e., stress or
anxiety) associated with the task [@sirois2013]. Furthermore, while also
conceptualised as a stable trait linked to specific personality factors
[@meng2024; @sirois2017c], these factors can shift in later life and may
interact with emerging cognitive or motivational changes to influence
procrastination behaviour in older adult [@oi2024; @donnellan2008].

Both apathy and procrastination have been linked to dysfunction in prefrontal
regions implicated in dementia [@friden2020; @zhang2010; @joseph2021;
@fahed2021]. Procrastination has been associated with alterations in brain
circuitry supporting cognitive control and emotional regulation. Research has
identified both structural and functional differences primarily in the
prefrontal cortex (PFC) and parahippocampal cortex [@zhang2019]. Specifically,
high procrastination is associated with decreased gray matter volume in the
dorsolateral PFC (dlPFC) and increased spontaneous activity in the ventromedial
PFC (vmPFC) [@liu2017]. Additionally, functional connectivity studies indicate
altered couplings between both the vmPFC and dlPFC, along with the vmPFC and
frontal gyri [@zhang2019]. Critically, these same networks are affected in
neurodegenerative conditions, including Alzheimer’s disease [@joseph2021].

These parallels raise the possibility that procrastination could serve as an
early marker of cognitive impairment or even a modifiable risk factor. Chronic
procrastination may exacerbate decline by limiting engagement in cognitively
stimulating activities such as physical exercise, problem-solving, and
goal-setting, which build cognitive resilience and reduce dementia risk
[@kelly2021; @mohammadibytamar2020; @chowdhary2022; @livingston2024]. Reduced
engagement may contribute to a cycle of cognitive disengagement, accelerating
decline.

@@

@@

To account for potential confounding, we controlled for demographic variables
with established associations with both cognitive function and procrastination
[@vanderweele2019]. These included measures of age, sex, and educational
attainment [@abner2012; @freedman2024; @yu2013]. Educational attainment was
classified into three categories: no formal education, GED (General Educational
Development)/high school diploma, and college/further education.

@@

All data analysis was carried out in R [@rcoreteam2025]. To model transitions in
cognitive states over time, we employed a first order discrete-time Markov
model, a class of stochastic processes that satisfy the Markov property
[@zhang2010], which can be formally expressed as:

@@

Our final analytic sample comprised $549$ respondents with the following age
distribution: 60 - 70 $(n = 186)$, 71 - 80 $(n = 203)$, 81 - 90 $(n = 142)$, and
90+ $(n = 18)$. Descriptive statistics for the full sample, as well as data
stratified by cognitive status (normative cognitive function, MCI, and
dementia), are presented in @tbl-descriptives. Both @fig-river and supplementary
figure S1 capture the unconditional transitions and transition probabilities
(respectively) between wave one and two (first transition), wave two and three
(second transition), and wave three and four (third transition), yielding a
total of $1,647$ observed transitions over time.

@@

# Table information
tbl_descriptives <- tibble::tribble(
  ~""                 , ~Full              , ~NC                , ~MCI              , ~Dementia        ,
  "Age (years)"       , "69.70 ± 7.58"    , "69.70 ± 7.55"    , "69.60 ± 7.55"   , "70.00 ± 9.89"  ,
  "Male"              , "38.43%<br>(n = 211)" , "39.82%<br>(n = 180)"  , "32.56%<br>(n = 28)" , "27.27%<br>(n = 3)" ,
  "Female"            , "61.57%<br>(n = 338)" , "60.18%<br>(n = 272)" , "67.44%<br>(n = 58)" , "72.73%<br>(n = 8)" ,
  "No degree"         , "16.24%<br>(n = 98)"  , "10.56%<br>(n = 47)"  , "39.53%<br>(n = 34)" , "63.54%<br>(n = 7)" ,
  "GED"               , "51.58%<br>(n = 279)" , "51.91%<br>(n = 231)" , "52.33%<br>(n = 45)" , "27.27%<br>(n = 3)" ,
  "Further education" , "32.29%<br>(n = 175)" , "37.53%<br>(n = 167)" , "8.14%<br>(n = 7)"   , "9.09%<br>(n = 1)"  ,
  "Apathy"            , "0.37 ± 0.63"     , "0.31 ± 0.59"     , "0.65 ± 0.76"    , "0.36 ± 0.51"   ,
  "Procrastination"   , "28.60 ± 12.00"   , "27.70 ± 11.30"   , "32.10 ± 13.80"  , "39.00 ± 16.00"
)

@@

# Outputting table
tbl_descriptives |>
  setNames(c(" ", "Full sample<br>(n = 549)", "NC<br>(n = 452)", "MCI<br>(n = 86)", "Dementia<br>(n = 11)")) |>
  tt(notes = list(caption,
  a = list(i = 10, j = 1, text = footnote))
  ) |>
   group_tt(i = list("Sex" = 2, "Education" = 4)) |>
   style_tt(j = 2:5, align = "c") |>
   style_tt(j = 1, align = "l") |>
   style_tt(i = c(2, 5), align = "l", bold = TRUE) |>
   # format_tt(escape = TRUE) |>
   format_tt(linebreak = "<br>") |>
   format_tt("notes", markdown = TRUE) |>
   style_tt(
     bootstrap_class = "table table-sm"
     )
```

@@

Results from the discrete-time Markov analysis, showed that all covariates
significantly influenced the likelihood of transitioning between cognitive
states (see @fig-markov). Notably, procrastination interacted significantly with
*age* to affect two key transitions: increasing the likelihood of transitioning
from normative cognitive function to MCI $(\text{OR} = 1.001; p < 0.001)$ and
decreasing the likelihood of reverting from MCI to normative cognitive function
$(\text{OR} = 0.999; p < 0.001)$. *Women* were significantly less likely than
men to transition from both normative cognitive function to dementia
$(\text{OR} = 0.068; p < 0.001)$ and from MCI to dementia
$(\text{OR} = 0.70; p < 0.001$).

For *education attainment*, individuals with a GED were less likely to
transition from normative cognitive function to either MCI
$(\text{OR} = 0.49; p < 0.001)$ or dementia $(\text{OR} = 0.29; p < 0.001)$ and
from MCI to dementia $(\text{OR} = 0.64; p < 0.001)$. They were also more likely
to back transition from MCI to normative cognitive function
$(\text{OR} = 2.18; p < 0.001)$. Those with a college level education or higher
demonstrated a significantly reduced likelihood of transitioning from normative
cognitive function to either MCI $(\text{OR} = 0.32; p < 0.001)$ or dementia
$(\text{OR} = 0.07; p < 0.001)$ and from MCI to dementia
$(\text{OR} = 0.22; p < 0.001)$. They were also more likely to back transition
from MCI to normative cognitive function $(\text{OR} = 3.13; p < 0.001)$.

Finally, higher levels of *apathy* were associated with an increased likelihood
of transitioning from normative cognitive function to both MCI
$(\text{OR} = 1.35; p = 0.005)$ and dementia $(\text{OR} = 1.36; p<0.001)$ and a
decreased likelihood of transitioning from MCI back to normative cognitive
function $(\text{OR} = 0.74; p = 0.003)$.

@@

Dementia poses a growing global health burden, with modifiable risk factors and
prodromal markers offering important targets for intervention [@livingston2024;
@livingston2020]. While apathy has been well established as a prodromal marker
and predictor of dementia [@donovan2014; @fresnais2023; @salem2023;
@vandalen2018], emerging evidence suggests that procrastination may also signal
early cognitive dysfunction [@friden2020; @zhang2019]. This study examined
whether procrastination, particularly in later life, could serve as a predictor
of cognitive decline. We assessed both cross-sectional differences in
procrastination across three cognitive groups (normative cognitive function, MCI
and dementia) and longitudinal associations with transitions in cognitive
status, to clarify whether procrastination functions as a novel and
age-sensitive marker of emerging cognitive impairment.

Our analysis revealed significant group differences in procrastination, with
individuals experiencing cognitive impairment, both MCI and dementia, reporting
higher procrastination scores than those with normative cognitive function.
These findings support the hypothesis that procrastination may be associated
with cognitive decline and aligns with emerging evidence linking procrastination
to cognitive dysfunction [@friden2020; @zhang2019]. Interestingly, while both
MCI and dementia groups exhibited elevated procrastination scores, no
significant difference emerged between these two groups. This suggests that
increases in procrastination may occur relatively early in the neurodegenerative
process, potentially preceding or paralleling the emergence of more overt
cognitive symptoms.

However, it should be noted that the number of participants classified as having
MCI $(n = 67)$ or dementia $(n = 27)$ was relatively small compared to those
with normative cognitive function $(n = 455)$. This imbalance in group sizes may
have limited the sensitivity of our comparisons and raises the possibility that
subtle differences between MCI and dementia groups were obscured by statistical
power constraints. Future research should aim to replicate these findings in
more evenly distributed samples to better determine whether procrastination
continues to increase across progressive stages of impairment or plateaus
following early decline.

@@

These findings imply that procrastination may serve dual roles, both as a marker
of early cognitive dysfunction and, potentially, as a behavioral impediment to
moving from MCI state to the normative cognitive state. One plausible pathway
involves the reinforcement of maladaptive behaviors such as reduced cognitive
engagement or reduced participation in cognitively protective activities, such
as physical exercise, social interaction or medical adherence [@hajek2025;
@kelly2021; @sirois2007; @stead2010]. In parallel, procrastination has been
associated with chronic stress and elevated cortisol levels [@sirois2023a],
which may accelerate hippocampal atrophy, β-amyloid plaque deposition, and brain
inflammation [@franks2021; @wallensten2023], further undermining cognitive
resilience. These behavioral and biological mechanisms align with
self-regulation theories suggesting that procrastination reflects executive
dysfunction [@rozental2014; @steel2007], an early feature of neurodegenerative
progression [@clark2012].

@@

Despite the insights offered by this study, several limitations should be
considered. Most notably, although our longitudinal Markov modelling approach
captured changes in cognitive status, a key constraint was the measurement of
procrastination at only a single time point (2020). As a result, we assumed
temporal stability in procrastination scores across time and used this measure
as both a retrospective and prospective proxy for procrastination scores.
However, this precluded analysis of within-person changes in this behavior over
time and whether such changes affect cognitive transitions. It should be noted
that this limitation is inherent to the HRS dataset rather than a methodological
oversight (Juster & Suzman, 1995). Future studies should prioritise repeated
assessments of procrastination to determine whether increases in procrastination
precedes, accompanies, or follows cognitive decline.

A further limitation concerns the classification of cognitive status. As our
analyses rely on the HRS, we are constrained by its operational definitions of
MCI and dementia [@crimmins2011], which are based on threshold scores on the
TICS. Although widely used [@langa2017; @davydow2015; @wei2018], these cut-offs
may not fully align with other clinical diagnostic criteria or
neuropsychological assessments, raising the possibility of misclassification.
Future research using more comprehensive cognitive batteries or clinical
diagnoses would help validate and refine the patterns observed here.

Additionally, negative emotional states (e.g., anxiety or depression) are
important psychological factors linked to both cognitive ageing and
procrastination. However, we were unable to adjust for anxiety in our models due
to substantial missingness in the HRS anxiety measure (approximately 61% missing
for our sample). Similarly, depression was not included as a covariate because
two CES-D items were used to construct our proxy measure of apathy, leaving a
non-standard six-item score that would not provide a valid index of depressive
symptoms. Future research using datasets with more complete emotional-health
assessments may help clarify the role of negative emotions in these
associations.

sMoreover, our use of a proxy measure of apathy may not have fully captured the
multidimensional nature of apathy. Future research should incorporate more
validated apathy assessments to strengthen the behavioral inferences drawn.
Additionally, as noted earlier, the relatively small number of individuals in
the MCI and dementia groups may have constrained statistical power for certain
comparisons and increased the potential for misclassification bias. Finally,
while the models adjusted for several key demographic and psychological
covariates, unmeasured confounding (undiagnosed medical conditions, medication
use, or sleep quality) could also have influenced cognitive outcomes. Future
research should seek to replicate and extend these findings using more
clinically diverse samples.

>>>>

---
title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
date: "2025-12-02"
author:
  - name: Cormac Monaghan
    url: https://c-monaghan.github.io/
    email: cormacmonaghan@protonmail.com
    orcid: 0000-0001-9012-3060
    affiliation:
      - id: 1
        name: Maynooth University
    attributes:
      corresponding: true
  - name: Rafael de Andrade Moral
    url: https://rafamoral.github.io/
    orcid: 0000-0002-0875-3563
    affiliation:
      - id: 1
        name: Maynooth University
  - name: Michelle Kelly
    url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
    orcid: 0000-0001-6159-6707
    affiliation:
      id: 2
      name: National College of Ireland
  - name: Joanna McHugh Power
    url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
    orcid: 0000-0002-7387-3107
    affiliation:
      id: 1
      name: Maynooth University
correspondence-prefix: "Corresponding author"
abstract: |
  Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential. While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction. We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} = 69.70;\sigma = 7.58)$. Cognitive function, procrastination, and a proxy measure of apathy were assessed. Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model. Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function. Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants. Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/C-Monaghan/crisp-clam>
keywords:
  - Dementia risk
  - Mild cognitive impairment
  - Procrastination
  - Apathy
  - Executive dysfunction
  - Aging
published: "This manuscript is a preprint and is currently under review at *Alzheimer's & Dementia: Diagnosis, Assessment & Disease Monitoring*"
code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
bibliography: references.bib
csl: csl/apa.csl
block-author: false
link-citations: true
editor:
  markdown:
    wrap: 80
---

@@

The Lancet Commission on Dementia identified fourteen causal risk factors
including smoking, hypertension, and diabetes, which account for approximately
45% of dementia cases worldwide [@livingston2024]. Prodromal markers, on the
other hand, reflect early disease processes and may signal limited opportunities
for prevention [@teipel2025]. One such prodromal marker is apathy, or the
significant loss of motivation, as distinct from depression and cognitive
impairment [@fresnais2023; @richard2012]. Apathy is prevalent in both MCI and
dementia subgroups [@donovan2014; @fresnais2023; @salem2023; @vandalen2018], and
is a robust predictor of progression from MCI to dementia [@ruthirakuhan2019;
@salem2023; @vandalen2018]. While apathy has been well-established as a
prodromal marker and risk factor for dementia progression, emerging evidence
suggests that procrastination may also relate to early cognitive decline.

Although apathy and procrastination can appear similar in that both lead to
reduced engagement in everyday activities, the underlying behavioural processes
are distinct. Apathy reflects reduced internal drive and emotional engagement,
impairing initiation of action [@fresnais2023]. In contrast, procrastination is
a self-regulatory failure defined by the voluntary delay of an important and
intended course of action despite expecting negative consequences [@sirois2013;
@klingsieck2013]. This distinction suggests that individuals who procrastinate
typically form intentions but fail to execute them in a timely manner. This
failure can often be understood through an affective framework, where delayed
action may stem from the avoidance of negative emotions (i.e., stress or
anxiety) associated with the task [@sirois2013]. Furthermore, while also
conceptualised as a stable trait linked to specific personality factors
[@meng2024], these factors can shift in later life and may interact with
emerging cognitive or motivational changes to influence procrastination
behaviour in older adult [@oi2024; @donnellan2008].

Both apathy and procrastination have been linked to dysfunction in prefrontal
regions implicated in dementia [@zhang2010; @joseph2021]. Procrastination has
been associated with alterations in brain circuitry supporting cognitive control
and emotional regulation. Research has identified both structural and functional
differences primarily in the prefrontal cortex (PFC) and parahippocampal cortex
[@zhang2019]. Specifically, high procrastination is associated with decreased
gray matter volume in the dorsolateral PFC (dlPFC) and increased spontaneous
activity in the ventromedial PFC (vmPFC). Additionally, functional connectivity
studies indicate altered couplings between both the vmPFC and dlPFC, along with
the vmPFC and frontal gyri [@zhang2019]. Critically, these same networks are
affected in neurodegenerative conditions, including Alzheimer’s disease
[@joseph2021].

These parallels raise the possibility that procrastination could serve as an
early marker of cognitive impairment or even a modifiable risk factor. Chronic
procrastination may exacerbate decline by limiting engagement in cognitively
stimulating activities such as physical exercise, problem-solving, and
goal-setting, which build cognitive resilience and reduce dementia risk
[@kelly2021; @livingston2024]. Reduced engagement may contribute to a cycle of
cognitive disengagement, accelerating decline.

@@

### Covariate: Depression

Depressive symptoms were included in our model based on their associations with
both procrastination [@klingsieck2013] and cognitive ageing [@livingston2020].
As mentioned, the HRS uses the CESD-8 as a measure of depressive symptomatology.
The scale consists of 8 items that are scored as either 0 (no) or 1 (yes), with
items 4 and 6 being reversed scored. Since we used two items from this scale as
proxies for apathy, we scored the remaining 6 items, giving a total score
ranging from 0 to 6, with higher scores indicating higher levels of depressive
symptoms. Across the waves, this 6-item scale showed reliable internal
consistency $(\alpha \in [0.81-0.83])$. An example of a question from the scale
includes “Much of the time during the past week, your sleep was restless”.

@@

To account for potential confounding, we controlled for demographic variables
with established associations with both cognitive function and procrastination
[@vanderweele2019]. These included measures of age, sex, and educational
attainment [@livingston2020; @steel2013]. Educational attainment was classified
into three categories: no formal education, GED (General Educational
Development)/high school diploma, and college/further education.

@@

All data analysis was carried out in R [@rcoreteam2025]. To model transitions in
cognitive states over time, we employed a first order discrete-time Markov
model, a class of stochastic processes that satisfy the Markov property, which
can be formally expressed as:

@@

Our final analytic sample comprised $549$ respondents with the following age
distribution: 60 - 70 $(n = 186)$, 71 - 80 $(n = 203)$, 81 - 90 $(n = 142)$, and
90+ $(n = 18)$. Descriptive statistics for the full sample, as well as data
stratified by cognitive status (normative cognitive function, MCI, and
dementia), are presented in @tbl-descriptives. Both @fig-river and Figure S1 in
the supplement capture the unconditional transitions and transition
probabilities (respectively) between wave one and two (first transition), wave
two and three (second transition), and wave three and four (third transition),
yielding a total of $1,647$ observed transitions over time.

@@

# Table information
tbl_descriptives <- tibble::tribble(
  ~""                 , ~Full                 , ~NC                   , ~MCI                 , ~Dementia           ,
  "Age (years)"       , "69.70 ± 7.58"       , "69.70 ± 7.55"       , "69.60 ± 7.55"      , "70.00 ± 9.89"     ,
  "Male"              , "38.43%<br>(n = 211)" , "39.82%<br>(n = 180)" , "32.56%<br>(n = 28)" , "27.27%<br>(n = 3)" ,
  "Female"            , "61.57%<br>(n = 338)" , "60.18%<br>(n = 272)" , "67.44%<br>(n = 58)" , "72.73%<br>(n = 8)" ,
  "No degree"         , "16.24%<br>(n = 98)"  , "10.56%<br>(n = 47)"  , "39.53%<br>(n = 34)" , "63.54%<br>(n = 7)" ,
  "GED"               , "51.58%<br>(n = 279)" , "51.91%<br>(n = 231)" , "52.33%<br>(n = 45)" , "27.27%<br>(n = 3)" ,
  "Further education" , "32.29%<br>(n = 175)" , "37.53%<br>(n = 167)" , "8.14%<br>(n = 7)"   , "9.09%<br>(n = 1)"  ,
  "Apathy"            , "0.37 ± 0.63"        , "0.31 ± 0.59"        , "0.65 ± 0.76"       , "0.36 ± 0.51"      ,
  "Depression"        , "0.98 ± 1.48"        , "0.88 ± 1.41"        , "1.41 ± 1.71"       , "1.09 ± 1.87"      ,
  "Procrastination"   , "28.60 ± 12.00"      , "27.70 ± 11.30"      , "32.10 ± 13.80"     , "39.00 ± 16.00"
)

@@

# Outputting table
tbl_descriptives |>
  setNames(c(" ", "Full sample<br>(n = 549)", "NC<br>(n = 452)", "MCI<br>(n = 86)", "Dementia<br>(n = 11)")) |>
  tt(notes = list(caption,
  a = list(i = 11, j = 1, text = footnote))
  ) |>
   group_tt(i = list("Sex" = 2, "Education" = 4)) |>
   style_tt(j = 2:5, align = "c") |>
   style_tt(j = 1, align = "l") |>
   style_tt(i = c(2, 5), align = "l", bold = TRUE) |>
   # format_tt(escape = TRUE) |>
   format_tt(linebreak = "<br>") |>
   format_tt("notes", markdown = TRUE) |>
   style_tt(
     bootstrap_class = "table table-sm"
     )
```

@@

Results from the discrete-time Markov analysis, showed that all covariates
significantly influenced the likelihood of transitioning between cognitive
states (see @fig-markov). Notably, *procrastination* interacted significantly
with *age* to affect two key transitions: increasing the likelihood of
transitioning from normative cognitive function to MCI
$(\text{OR} = 1.001; p < 0.001)$ and decreasing the likelihood of reverting from
MCI to normative cognitive function $(\text{OR} = 0.999; p < 0.001)$. *Women*
were significantly less likely than men to transition from both normative
cognitive function to dementia $(\text{OR} = 0.67; p < 0.001)$ and from MCI to
dementia $(\text{OR} = 0.71; p < 0.001$).

For *education attainment*, individuals with a GED were less likely to
transition from normative cognitive function to either MCI
$(\text{OR} = 0.46; p < 0.001)$ or dementia $(\text{OR} = 0.29; p < 0.001)$ and
from MCI to dementia $(\text{OR} = 0.64; p < 0.001)$. They were also more likely
to back transition from MCI to normative cognitive function
$(\text{OR} = 2.24; p < 0.001)$. Those with a college level education or higher
demonstrated a significantly reduced likelihood of transitioning from normative
cognitive function to either MCI $(\text{OR} = 0.31; p < 0.001)$ or dementia
$(\text{OR} = 0.07; p < 0.001)$ and from MCI to dementia
$(\text{OR} = 0.22; p < 0.001)$. They were also more likely to back transition
from MCI to normative cognitive function $(\text{OR} = 3.23; p < 0.001)$.

Higher levels of *depression* were associated with an increased likelihood of
transitioning from normative cognitive function to MCI
$(\text{OR} = 1.14; p = 0.013)$ and a reduced likelihood of transitioning back
from MCI to normal cognition $\text{OR} = 0.88; p = 0.013)$. Finally, higher
levels of *apathy* were associated with an increased likelihood of transitioning
from normative cognitive function to dementia $(\text{OR} = 1.20; p < 0.001)$

@@

Dementia poses a growing global health burden, with modifiable risk factors and
prodromal markers offering important targets for intervention [@livingston2024;
@livingston2020]. While apathy has been well established as a prodromal marker
and predictor of dementia [@donovan2014; @fresnais2023; @salem2023;
@vandalen2018], emerging evidence suggests that procrastination may also signal
early cognitive dysfunction [@joseph2021; @zhang2019]. This study examined
whether procrastination, particularly in later life, could serve as a predictor
of cognitive decline. We assessed both cross-sectional differences in
procrastination across three cognitive groups (normative cognitive function, MCI
and dementia) and longitudinal associations with transitions in cognitive
status, to clarify whether procrastination functions as a novel and
age-sensitive marker of emerging cognitive impairment.

Our analysis revealed significant group differences in procrastination, with
individuals experiencing cognitive impairment, both MCI and dementia, reporting
higher procrastination scores than those with normative cognitive function.
These findings support the hypothesis that procrastination may be associated
with cognitive decline and aligns with emerging evidence linking procrastination
to cognitive dysfunction [@joseph2021; @zhang2019]. Interestingly, while both
MCI and dementia groups exhibited elevated procrastination scores, no
significant difference emerged between these two groups. This suggests that
increases in procrastination may occur relatively early in the neurodegenerative
process, potentially preceding or paralleling the emergence of more overt
cognitive symptoms.

However, it should be noted that the number of participants classified as having
MCI $(n = 67)$ or dementia $(n = 27)$ was relatively small compared to those
with normative cognitive function $(n = 455)$. This imbalance in group sizes may
have limited the sensitivity of our comparisons and raises the possibility that
subtle differences between MCI and dementia groups were obscured by statistical
power constraints. In addition, our analytic sample contained a higher
proportion of women than men $(61.57\% \text{female})$ reflecting the
demographic composition of the available HRS respondents who met eligibility
criteria. Although this imbalance was not the result of a sampling decision, it
may influence the generalisability of our findings, as women in older cohorts
differ from men in both dementia risk profiles and health-seeking behaviours
[@berve2025]. Future research should aim to replicate these findings in more
evenly distributed samples to better determine whether these results are
consistent across men and women and whether procrastination continues to
increase across progressive stages of impairment or plateaus following early
decline.

@@

These findings imply that procrastination may serve dual roles, both as a marker
of early cognitive dysfunction and, potentially, as a behavioral impediment to
moving from MCI state to the normative cognitive state. One plausible pathway
involves the reinforcement of maladaptive behaviors such as reduced cognitive
engagement or reduced participation in cognitively protective activities, such
as physical exercise, social interaction or medical adherence [@hajek2025;
@kelly2021; @sirois2007; @stead2010]. In parallel, procrastination has been
associated with chronic stress and elevated cortisol levels [@sirois2023a],
which may accelerate hippocampal atrophy, β-amyloid plaque deposition, and brain
inflammation [@franks2021; @wallensten2023], further undermining cognitive
resilience. These behavioral and biological mechanisms align with
self-regulation theories suggesting that procrastination reflects executive
dysfunction [@rozental2014], an early feature of neurodegenerative progression
[@clark2012]. At the same time, these results align with both the affective and
trait-based perspectives on procrastination. Higher procrastination in
individuals with cognitive impairment may partly reflect affective dysregulation
(e.g., anxiety or stress-related avoidance) [@sirois2013] or age-related shifts
in stable behavioural tendencies that accompany neurodegenerative change
[@oi2024]. While our study cannot differentiate between these frameworks, our
results highlight that procrastination is a multifaceted construct encompassing
motivational, affective, and self-regulatory dimensions.

@@

Despite the insights offered by this study, several limitations should be
considered. Most notably, although our longitudinal Markov modelling approach
captured changes in cognitive status, a key constraint was the measurement of
procrastination at only a single time point (2020). As a result, we assumed
temporal stability in procrastination scores across time and used this measure
as both a retrospective and prospective proxy for procrastination scores.
However, this precluded analysis of within-person changes in this behavior over
time and whether such changes affect cognitive transitions. It should be noted
that this limitation is inherent to the HRS dataset rather than a methodological
oversight [@juster1995]. Future studies should prioritise repeated assessments
of procrastination to determine whether increases in procrastination precedes,
accompanies, or follows cognitive decline.

A further limitation concerns the classification of cognitive status. As our
analyses rely on the HRS, we are constrained by its operational definitions of
MCI and dementia [@crimmins2011], which are based on threshold scores on the
TICS [@brandt1988; @fong2009]. Developed as a telephone-based analogue of the
Mini-Mental State Examination, the TICS is a widely adopted and well-validated
cognitive screener [@fong2009]. However, like all brief screening measures,
TICS-based classifications may not fully align with more detailed clinical
assessments, raising the possibility of diagnostic misclassification. Future
research using more comprehensive cognitive batteries or clinical diagnoses
would help validate and refine the patterns observed here.

Additionally, negative emotional states (e.g., anxiety) are important
psychological factors linked to both cognitive ageing and procrastination
[@livingston2020; @sirois2013]. However, we were unable to robustly adjust for
anxiety in our models due to substantial systematic missingness in the HRS
anxiety measure (see Figure S2 and Figure S3 in the supplement). Moreover, our
use of a proxy measure of apathy may not have fully captured the
multidimensional nature of apathy. Additionally, it meant that only six of the
original eight CESD-8 items were used for scoring depression, possibly
introducing measurement error. Future research should incorporate more validated
apathy assessments to strengthen the behavioral inferences drawn.

Finally, while the models adjusted for several key demographic and psychological
covariates, unmeasured confounding (undiagnosed medical conditions, medication
use, or sleep quality) could also have influenced cognitive outcomes. Future
research should seek to replicate and extend these findings using more
clinically diverse samples.
